---
title: "Lab 3"
author: "vaibhavi"
date: "January 27, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Movies Dataset: More Analysis!

### Load the data.
```{r}
movies=read.csv("movies.csv")
ratings=read.csv("ratings.csv")
```


### a) How many unique genres of movies are there?
There are 19 overall genres, if we exclude the "(no genres listed)" label applied in the genre column.
Used strsplit to create lists within each row.
Used unlist to break all the lists within each row into individual genres.
```{r}
genres = strsplit(as.character(movies$genre), "[|]")
allgenres = unlist(genres)
uniquegenres= unique(allgenres)
uniquegenres

```



### b) Of the unique genres, what are the 5 most popular?
Top 5 are Drama, Comedy,Thriller, Action, Romance
```{r}
#uniquegenres
#names(uniquegenres) is NULL, because uniquegenres is a Vector

#genremovies = data.frame(genre = uniquegenres)
#genremovies$genre = uniquegenres

#genremovies$movienum=lapply(uniquegenres, function(x)  length(grep(paste("\\<", x, "\\>", sep=""), ratings$genre)) );
#genremovies
#find_instances = function(x) { length(grep(paste("\\<", x, "\\>", sep="")), ratings$genre) }
#find_instances(x="Drama")

freqtable =table(allgenres)
freqtable[order(table(allgenres), decreasing=TRUE)]


```



### c) Distribution of Number of ratings per user. Are there "super raters" who review much more than others?

```{r}
#users= ratings$userId
#hist(users, breaks=length(users)-1)

ratingsuser=data.frame(table(ratings$userId))
#dim(ratingsuser)
colnames(ratingsuser)=c("UserId", "FilmsRated")

#To get raters in decreasing order
#Need to sort so that we get a better idea of distribution
#ratingsuser=ratingsuser[order(ratingsuser$FilmsRated, decreasing="TRUE"),]
#ratingsuser

#Keeping only the Films Rated, in Descending Order
ratedvalues=ratingsuser[order(ratingsuser$FilmsRated, decreasing="TRUE"), "FilmsRated"] 


 hist(ratedvalues, breaks=length(ratedvalues)-1, main="Histogram of Number of Ratings per user", xlab="Number of films rated", ylab="Frequency")
 
#Most prolific User Id is 547 
cat("Details of Most Prolific User \n")
ratingsuser[which(ratingsuser$FilmsRated == max(ratingsuser$FilmsRated)), ]

avgcount= mean(ratingsuser$FilmsRated)
cat(paste("Average films rated :" , avgcount))
sdcount= sd(ratingsuser$FilmsRated)
upperCount= avgcount + 3*sdcount


#plot(ratingsuser$UserId, ratingsuser$FilmsRated, pch=4)
#abline(h=avgcount, lwd=2, col="red")

```



### d) Plot number of movies rated (by year) of the most prolific rater. What does this trend look like?
```{r}
awesomeratings=ratings[which(ratings$userId == 547),]
#awesomeratings

#We need to make a table which keeps only the year, and the corresponding frequency
yeartable =table(awesomeratings$year)
yearlyratings= data.frame(yeartable)
colnames(yearlyratings) = c("Year","UserRated")

#plot yearly ratings of User Id 547 
plot(yearlyratings, ylim=c(0,300), main="User 547's Rated Films vs Total(Blue)", xlab="Year", ylab="Number of Films")
abline(h=mean(yearlyratings$UserRated), col="red", lwd=2)

#Add points for yearly total movies in dataset
yearlymovies = data.frame(table(movies$year))
colnames(yearlymovies) = c("Year", "TotalMovies")
points(yearlymovies, col="blue")

#points(yeartable, col="green")
```

